---
title: "ds4eeb Final Project"
author: "Jake Metzger, Niki Shridar, Haleigh Yang"
date: today
format: pdf
editor: visual
---

In this script we are going to analyzing data from the Kroeker Labs annual surveys of rocky subtidal reef communites near Sitka, Alaska. Over the past few years, we have seen the transition of these reefs from kelp forests to urchin barrens, and the loss of abalone populations. So the motivating questions here are:

Quantitatively, can we see these trends in our data?

-And-

What is the relationship between urchin and abalone populations?

Our target audience is the general public and scientific community as we plan to share these results (after some refinement) with the community of Sitka, the scientific community, and, perhaps, resource managers.

Note: we are only using data from summer surveys, as winter data may obscure our results due to the seasonal fluctuations in algal and fauna communities. So, let's start loading some libraries and read in the data:\
\
\

## Loading libraries

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(janitor)
library(here)
library(readr)
library(purrr)
library(scales)
library(lubridate)
library(RColorBrewer)
library(brms) # for statistics
library(dagitty) # for loading DAGs
library(patchwork)
library(tidybayes)
```

## Reading and plotting the data

```{r}
#| message: false
#| warning: false

swath_density <- read_csv(here("data/swath_density.csv")) %>% 
  mutate(
    transect = as.character(transect),
    transect = as.numeric(case_when(
      transect == "upcoast"   ~ "1",
      transect == "downcoast" ~ "2",
      TRUE ~ transect
    )),
    zone = recode(zone, "perm" = "deep")
  ) %>% 
  filter(zone != "mid", season == "summer") %>%  # filter out odd zone code, and only include summer data
  select(-season) #remove season column as only using summer data

swath_size_freq <- read_csv(here("data/swath_size_freq.csv")) %>% 
  mutate(
    transect = as.character(transect),
    transect = as.numeric(case_when(
      transect == "upcoast"   ~ "1",
      transect == "downcoast" ~ "2",
      TRUE ~ transect
    )),
    zone = recode(zone, "perm" = "deep")
  ) %>% 
  filter(zone != "mid", segment != "all", season == "summer") %>%  # filter out odd zone and segment codes, and only include summer data
  select(-season)

# make key of all real sampling events to fill zeros later
sampling_key <- swath_size_freq %>%
  distinct(year, site, zone, transect, segment)

```

### Abalone biomass data

Using the equation W = 0.0000578L\^3.2 from Zhang et al 2007 to convert shell length (L in mm) into body weight (W in g). The two parameter values were the means estimated by Breen & Adkins (1982).

Citation: Zhang, Z., Campbell, A., & Lessard, J. (2007). Modeling northern abalone, Haliotis kamtschatkana, population stock and recruitment in British Columbia. Journal of Shellfish Research, 26(4), 1099-1107.

Pull in swath size frequency data, filter for abalone, sum at the segment level, calculate average biomass density (g/m\^2) at the transect level for plotting.

```{r}
ab_biodensity_segment <- swath_size_freq %>% 
  filter(classcode == "halkam") %>% #filter for abalone
  mutate(biomass_g = 0.0000578 * size_or_stipe_count^3.2) %>%
  group_by(year, site, zone, transect, segment) %>%
  summarise(
    biomass_density = sum(biomass_g, na.rm = TRUE) / 20,
    .groups = "drop"
  ) %>%
  right_join(sampling_key, by = c("year", "site", "zone", "transect", "segment")) %>%
  mutate(biomass_density = replace_na(biomass_density, 0))

ab_biodensity_transect <- ab_biodensity_segment%>%
  group_by(year, site, zone, transect) %>%
  summarise(mean_biomass_density = mean(biomass_density, na.rm = TRUE), .groups = "drop")
```

### Plotting Abalone

```{r}
ggplot(ab_biodensity_transect, 
       aes(x = year, y = mean_biomass_density, color = zone)) +
  geom_point() +
  geom_smooth() 
```

### Macro biomass density

**Constants for macro from Lauren Bell's paper:** macro_wet_slope \<- 734.9 macro_dry_slope \<- 0.103 survey_area_m2 \<- 30 \* 2 \# 60m² segment_area_m2 \<- survey_area_m2 / 3 \# 20m²

```{r}

macro_wet_slope <- 734.9 
macro_dry_slope <- 0.103 
#transect_area_m2 <- 20 * 3 # 60m² 
#segment_area_m2 <- 10 * 2 # 20m²

macro_biodensity_segment <- swath_size_freq %>% 
  filter(classcode == "macpyrad") %>%
  mutate(dry_mass_g_plant = size_or_stipe_count * macro_wet_slope * macro_dry_slope) %>%
  group_by(year, site, zone, transect, segment) %>%
  summarise(
    biomass_density = (sum(dry_mass_g_plant, na.rm = TRUE) / 20),
    .groups = "drop"
  ) %>%
  right_join(sampling_key, by = c("year", "site", "zone", "transect", "segment")) %>%
  mutate(biomass_density = replace_na(biomass_density, 0))
  
macro_biodensity_transect <- macro_biodensity_segment %>%   
  group_by(year, site, zone, transect) %>%
  summarise(mean_biomass_density = mean(biomass_density, na.rm = TRUE), .groups = "drop") 
```

### Plotting Macrocystis

```{r}
ggplot(macro_biodensity_transect, 
       aes(x = year, y = mean_biomass_density, color = zone)) +
  geom_smooth(alpha = 0.6) +
  geom_point(size = 2, alpha = 0.7) +
  #facet_wrap(~ site, scales = "free_y") +
  labs(
    x = "Year",
    y = "Frond Standing Crop (g/m²)",
    title = "Kelp Dry Mass Across Years by Site") +
  theme_classic()
```

### Urchin Data

converting test diameter to biomass

Conversion factors from Lochead, J., W. Hajas, and Don Leus. Calculation of mean abundance in the Red Urchin Analysis Program and Green Urchin Analysis Program. Fisheries and Oceans Canada, Science Branch, Pacific Region, Pacific Biological Station, 2015.

Reds (and using for purps): W = 0.0010169 × T\^2.7787 Greens: W = 11.2181 × (T/30)\^2.754085

```{r}

custom_colors <- c("Green Urchin" = "#90ee90",
                   "Red Urchin" = "red")
              

urch_biodensity_segment_species <- swath_size_freq %>%
  filter(classcode %in% c("strdro","strfraad")) %>%
  filter(size_or_stipe_count < 150) %>% #removing outliers
  mutate(
    common_name = case_when(classcode == "strdro" ~ "Green Urchin",
                                 classcode == "strfraad" ~ "Red Urchin"),
    site = case_when(site == "Breast" ~ "BREAST", #cleaning
                          site == "Harris" ~ "HARRIS",
                          .default = site),
    biomass_density = case_when(
      classcode == "strdro" ~ 11.2181 * (size_or_stipe_count / 30)^2.754085,
      classcode %in% c("strfraad") ~ 0.0010169 * size_or_stipe_count^2.7787,
    TRUE ~ NA_real_
  )) %>% 
  # summing at the segment level by species and divide by area sampled
  group_by(year, site, zone, transect, segment, common_name) %>%
  summarise(biomass_density= sum(biomass_density, na.rm = TRUE) / first(area_samp),
            .groups = "drop") %>% 
  right_join(sampling_key, by = c("year", "site", "zone", "transect", "segment")) %>%
  mutate(biomass_density = replace_na(biomass_density, 0))

#1: Sum both species within each segment
urch_biodensity_segment <- urch_biodensity_segment_species %>%
  group_by(year, site, zone, transect, segment) %>%
  summarise(biomass_density = sum(biomass_density, na.rm = TRUE), .groups = "drop")

#2: Average segments up to transect level
urch_biodensity_transect <- urch_biodensity_segment %>%
  group_by(year, site, zone, transect) %>%
  summarise(mean_biomass_density = mean(biomass_density, na.rm = TRUE), .groups = "drop")
```

### Plotting Urchins

```{r}
       
ggplot(urch_biodensity_transect, aes(x = year, y = mean_biomass_density, color = zone)) +
  geom_point() +
  geom_smooth() +
  theme_classic() 
```

### Understory Algae Data

Understory algae is simply density (n/m^2), summed across subcanopy taxa as we do not have biomass conversions for most species.

```{r}
under_density_segment <- swath_density %>% #reading from density df as we are not converting to biomass
  filter(class == "algae", !classcode %in% c("macpyrad", "macpyrjv", "unkalg")) %>% #Remove Macrocystis data and unknow algae
  group_by(year, site, zone, transect, segment) %>%
  summarise(
    under_density = (sum(density, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  right_join(sampling_key, by = c("year", "site", "zone", "transect", "segment")) %>%
  mutate(under_density = replace_na(under_density, 0)) 
  
under_density_transect <- under_density_segment %>% 
group_by(year, site, zone, transect) %>%
  summarise(mean_under_density = mean(under_density, na.rm = TRUE), .groups = "drop") 
```

### Plotting Understory

```{r}
ggplot(under_density_transect, 
       aes(x = year, y = mean_under_density, color = zone)) +
  geom_smooth() +
  geom_point() +
  labs(
    x = "Year",
    y = "Density (n/m²)",
    title = "Understory Algal Density Across Years by Zone") +
  theme_classic()
```

### Pycno Data

pycno biomass conversion = exp(-3.9989)*(length.cm^3.133)

Lee, L. C., J. C. Watson, R. Trebilco, and A. K. Salomon. 2016. Indirect effects and prey behavior mediate interactions between an endangered prey and recovering predator. Ecosphere 7(12):e01604. 10.1002/ecs2.1604

```{r}
#Pycno biomass calculations

pyc_biodensity_segment <- swath_size_freq %>%
  filter(classcode == "pychel") %>%
  mutate(biomass_g = exp(-3.9989) * (size_or_stipe_count^3.133)) %>%
  group_by(year, site, zone, transect, segment) %>%
  summarise(
    biomass_density = sum(biomass_g, na.rm = TRUE) / 20,
    .groups = "drop"
  ) %>%
  mutate(biomass_density = replace_na(biomass_density, 0))%>%
  right_join(sampling_key, by = c("year", "site", "zone", "transect", "segment")) %>% 
  mutate(biomass_density = replace_na(biomass_density, 0))

pyc_biodensity_transect <- pyc_biodensity_segment %>% 
  group_by(year, site, zone, transect) %>%
  summarise(mean_biomass_density = mean(biomass_density, na.rm = TRUE), .groups = "drop")
```

### Plotting Pycno Data

```{r}
ggplot(pyc_biodensity_transect, 
       aes(x = year, y = mean_biomass_density, color = zone)) +
  geom_smooth() +
  geom_point() +
  labs(
    x = "Year",
    y = "Biomas density (g/m^2)",
    title = "Pycnopodia Biomass Density Across Years by Zone") +
  theme_classic()
```

## Putting it all together

### Combining all data

```{r}
#at segment level for model
combined_segments <- bind_rows(
  ab_biodensity_segment %>%
    mutate(taxa = "Abalone"),
  macro_biodensity_segment %>% 
    mutate(taxa = "Macrocystis"),
  pyc_biodensity_segment %>% 
    mutate(taxa = "Pycnopodia"),
  under_density_segment %>% 
    rename(biomass_density = under_density) %>% 
  mutate(taxa = "Understory"),
  urch_biodensity_segment %>% 
    mutate(taxa = "Urchins"))

#At transect level for plotting
combined_transects <- bind_rows(
  ab_biodensity_transect %>% 
    mutate(taxa = "Abalone"),
  macro_biodensity_transect %>% 
    mutate(taxa = "Macrocystis"),
  urch_biodensity_transect %>% 
    mutate(taxa = "Urchins"), 
  under_density_transect %>% 
    rename(mean_biomass_density = mean_under_density) %>% 
    mutate(taxa = "Understory Algae"), 
  pyc_biodensity_transect %>% 
    mutate(taxa = "Pycnopodia")
)
```

```{r}
#Normalizing data, max = 1, 0 = 0
combined_normal0 <- combined_transects %>%
  group_by(taxa) %>%
  mutate(mean_scaled = mean_biomass_density / max(mean_biomass_density, na.rm = TRUE)) %>%
  ungroup()
```
### Plotting taxa over time
```{r}
taxa_colors <- c(
  "Urchins"       = "mediumorchid4",
  "Abalone"          = "#FFB6C1",
  "Understory Algae" = "yellow3",
  "Macrocystis"      = "olivedrab3", 
  "Pycnopodia"  = "lightblue"
)

ggplot(combined_normal0, aes(x = year, y = mean_scaled, color = taxa, fill = taxa)) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 5), se = TRUE, alpha = 0.15) +
  scale_color_manual(values = taxa_colors) +
  scale_fill_manual(values = taxa_colors) +
  coord_cartesian(ylim = c(0, NA)) +
  labs(x = "Year", y = "Normalized %", color = "Taxa", fill = "Taxa") +
  theme_classic()
```

### Plotting urchins against other taxa

### Reformatting data for model

```{r}
# pivot to wide so each taxa becomes its own column,
# allowing us to use Urchins as the x variable in the plot
combined_wide_plot <- combined_transects %>%
  group_by(year, site, zone, transect, taxa) %>%
  summarise(mean_biomass_density = mean(mean_biomass_density, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = taxa,
              values_from = mean_biomass_density)

# pivot back to long but exclude Urchins — it stays as its own column (the x axis)
combined_long_plot <- combined_wide_plot %>%
  pivot_longer(cols = c("Abalone", "Macrocystis", "Understory Algae", "Pycnopodia"),
               names_to = "taxa",
               values_to = "mean_biomass_density")
```

### Plotting Urchins V Other Taxa

```{r}
ggplot(combined_long_plot, aes(x = Urchins, y = mean_biomass_density, color = taxa, fill = taxa)) +
  geom_point(alpha = 0.4, size = 2) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 5), se = TRUE, alpha = 0.15) +
  scale_x_continuous(
    trans = "log1p",
    breaks = c(0, 1, 10, 100, 1000),
    labels = c("0", "1", "10", "100", "1000")
  ) +
  geom_vline(xintercept = c(1, 10, 100, 1000), linetype = "dashed", color = "grey60") +
  facet_wrap(~ factor(taxa, levels = c("Macrocystis", "Understory Algae", "Abalone", "Pycnopodia")),
             scales = "free_y",
             labeller = as_labeller(c(
               "Macrocystis"      = "Macrocystis (g/m²)",
               "Understory Algae" = "Understory Algae (n/m²)",
               "Abalone"          = "Abalone (g/m²)",
               "Pycnopodia"       = "Pycnopodia (g/m²)"
             ))) +
  scale_color_manual(values = taxa_colors) +
  scale_fill_manual(values = taxa_colors) +
  coord_cartesian(ylim = c(0, NA)) +
  labs(x = "Urchin Biomass Density (g/m²), log(x + 1) scale)",
       y = "Density") +
  theme_classic(base_size = 13) +
  theme(
    legend.position = "none",
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    panel.spacing = unit(1.5, "lines")
  )
```

### Our DAG

```{r}
DAG <- dagitty('dag {
bb="0,0,1,1"
Pycnopodia [adjusted,pos="0.100,0.500"]
Urchins [exposure,pos="0.300,0.500"]
Macrocystis [pos="0.500,0.350"]
UnderstoryAlgae [pos="0.500,0.650"]
Abalone [outcome,pos="0.700,0.500"]
Macrocystis -> Abalone
Pycnopodia -> Abalone [pos="0.400,0.300"]
Pycnopodia -> Urchins
UnderstoryAlgae -> Abalone
Urchins -> Macrocystis
Urchins -> UnderstoryAlgae
}')
plot(DAG)
```

### The BRM Model

```{r}
#| warning: false

# ------------------------------------------------------------
# 1. Build modeling dataframe
# ------------------------------------------------------------

# Pivot wide so each taxa is its own column
combined_wide_model <- combined_segments %>%
  pivot_wider(names_from = taxa,
              values_from = biomass_density) %>%
  clean_names() %>%  # janitor::clean_names() to handle spaces/caps
  mutate( site = factor(site),
    zone = factor(zone),
    transect = factor(transect),
    segment = factor(segment),
    year = factor(year),
    macrocystis = log1p(macrocystis),
    understory = log1p(understory),
    pycnopodia = log1p(pycnopodia),
    urchins = log1p(urchins)
  )

# Check structure
head(combined_wide_model)
```

```{r}
# ------------------------------------------------------------
# 2. Define model formulas (trophic cascade DAG)
# ------------------------------------------------------------

# Pycnopodia → Urchins
f_urchins <- bf(
  urchins ~ pycnopodia +
    (1 | year) +
    (1 | site/zone/transect/segment),
  family = hurdle_gamma()
)

# Urchins → Macrocystis
f_macrocystis <- bf(
  macrocystis ~ urchins +
    (1 | year) +
    (1 | site/zone/transect/segment),
  family = hurdle_gamma()
)

# Urchins → Understory algae
f_understory <- bf(
  understory ~ urchins +
    (1 | year) +
    (1 | site/zone/transect/segment),
  family = hurdle_gamma()
)

# Kelp + predators → Abalone
f_abalone <- bf(
  abalone ~ macrocystis + understory + pycnopodia +
    (1 | year) +
    (1 | site/zone/transect/segment),
  family = hurdle_gamma()
)

# ------------------------------------------------------------
# 3. Priors
# ------------------------------------------------------------

priors <- c(
  prior(normal(0,1), class = "b"),
  prior(normal(0,2), class = "Intercept"),

  prior(exponential(1), class = "shape", resp = "urchins"),
  prior(exponential(1), class = "shape", resp = "macrocystis"),
  prior(exponential(1), class = "shape", resp = "understory"),
  prior(exponential(1), class = "shape", resp = "abalone")
)

# ------------------------------------------------------------
# 4. Fit multivariate trophic cascade model
# ------------------------------------------------------------

cascade_model <- brm(
  f_urchins +
    f_macrocystis +
    f_understory +
    f_abalone +
    set_rescor(FALSE),

  data = combined_wide_model,
  prior = priors,

  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,

  control = list(adapt_delta = 0.95)
)
```

```{r}
# ------------------------------------------------------------
# 5. Diagnostics
# ------------------------------------------------------------

summary(cascade_model)


plot(cascade_model)
```
